---
title: "0X_Analysis_Graft_vs_Host_disease"
format: html
editor: visual
---

# Graft vs. host disease in relation to Survival

Dataset: *Bone Marrow Transplantation: Children*

We want to look at the relationship between relapse and secondary transplants in patients and how it affects their odds of survival.

Variables used:

| Variable | Description | Value | Missing values |
|:--:|:--:|:--:|----|
| `aGvHDIIIIV` | Development of acute graft versus host disease stage III or IV | Yes, No | No, kind of |
| `extcGvHD` | Development of extensive chronic graft versus host disease | Yes, No | Yes |
| `time_to_aGvHD_III_IV` | Time to development of acute graft versus host disease stage III or IV | Integer, time in days | No |
| X? |  |  |  |
| X? |  |  |  |

: Table 1: The data used for this analysis is from `03_dat_aug` ? using these variables: `aGvHDIIIIV`, `extcGvHD`, `time_to_aGvHD_III_IV`, X & X.

## Data Wrangling

### Initiate Libraries

We start by loading the relevant libraries.

```{r}
library(tidyverse)
#library("table1")

```

We then load a subset of the cleaned and augmented data frame as a tibble.

```{r}

bm_data <- read_tsv(file = "data/03_dat_aug.tsv",
                            show_col_types = FALSE)

```

It contains 186 values and follows the guidelines for tidy data, with:

-   Each variable having its own column.

-   Each observation having its own row.

-   Each value having its own cell.

Some variables used in this analysis contains missing values or very high values, see table 1 for more details. The high values correspond with other data.

The full tibble can be seen here.

```{r}
bm_data
```

## Data Analysis

### Development of Acute Graft

We look at the time it takes to develop acute graft and SOMETHING WITH VERSUS HOST DISEASE STAGE III AND IV.

We do this by creating a subset using X variables to look into the development of acute graft vs. host disease stage III & IV.

```{r}
acute_graft_data <- bm_data |>
  select(aGvHDIIIIV,
         time_to_aGvHD_III_IV)

acute_graft_data
```

Further look into values.

```{r}
acute_graft_data |>
  filter(aGvHDIIIIV == "Yes") |>
  summarise()
```

We then see that entries with no development of acute graft has exstremly high values of 10\^6 days, where cases of development falls in
