---
title: "0X_Relapse_and_second_transplant_vs_survival_of_patients"
format: html
editor: visual
---

# Relapse & Second Transplant vs. Survival of Patients

We want to look at the relationship between relapse and secondary transplants in patients and how it affects their odds of survival.

To Do this, we look at how many relapsed, needed a second transplant, and how many of them survived and when they where observed. This is done using 4 variables taken from xxx, as described in table 1, and then performing an analysis of the relationship between them.

| Variable | Description | Value |
|:---------------------:|:----------------------------------------------------------------:|:-----------------------:|
| `Txpostrelapse` | Second Bone Marrow transplantation after relapse | Yes (1) or No (0) |
| `Relapse` | Reoccurance of disease | Yes (1) or No (0) |
| `survival_time` | Time of Observation (Alive) or time of event in days (Dead) | Yes (1) or No (0) |
| `survival_status` | Survival status | Yes (1) or No (0) |

: Table 1: The data used for this analysis is xxx using these variables:`Txpostrelapse`, `Relapse`, `survival_time`& `survival_status`.

## Initiate Libraries

We start by loading the relevant libraries.

```{r}
library(tidyverse) 
```

## Data Wrangling

First we load the xxx data as a tibble.

```{r}

bm_data <- read_tsv("data/03_dat_aug.tsv.gz")



```

```{r}
bm_data

```

compare data

```{r}
ggplot(data = bm_data,
       mapping = aes(x = HLAmatch,
                     group = survival_status)) +
  geom_histogram()

  
```

Old plots not edited

```{r}
#box plot with stradified expression levels
ggplot(data = bm_data,
       mapping = aes(x = HLAmatch,
                     y = survival_time,
                     fill = HLAmatch)) +
  #boxplot
  geom_boxplot(alpha = 0.45)
```

100000 -\> used as placeholder for nothing in some columns

Augment (PCA): time_to_aGvHD_III_IV,

lm - function,

Gendermatch -\> ung men prefered, sensitation

HLAmatch

**subdataset**

group together to see if dead or observation

-   survival_time

Time of observation (if alive) or time to event (if dead) in days

-   survival_status

Survival status (0 - alive, 1 - dead)

psudosample number

first or second transplant
