---
title: "CD34+ and CD3 analysis"
author: "inma"
format: html
editor: visual
---

# 1. LOAD LIBRARIES

```{r}

library(tidyverse)
library(ggplot2)
library(dplyr)
```

# 2. LOAD DATA

```{r}
data_clean <-  read_tsv('data/02_dat_clean.tsv.gz')
```

# 3. ANALYSIS

## 3.1. CD34+ ANALYSIS 

**Description**:

In this plot, the relationship between CD34+ cell dose and post-transplant survival was analyzed. Patients predominantly received low CD34+ doses (0-25 cells/kg), with survival times ranging from approximately 500 - 3500 days. Linear regression revealed a near-flat, slightly positive slope, indicating minimal influence of CD34+ dose on survival. The widening confidence interval at higher doses reflects increased uncertainty due to fewer data points. Overall, CD34+ dose does not appear to predict survival in this patient group, with both low and high dose recipients showing highly variable outcomes

```{r}

# Filter only the survival group 
group_surv <- function(data) {
  data_clean |>
    filter(survival_status == '0')
}

# Plot using the fuction 
ggplot(group_surv(), 
       aes(x = CD34kgx10d6, 
           y = survival_time)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = TRUE) + 
  labs(title = 'CD34+ dose vs Survival Days',
       x = 'CD34+ cells/kg', 
       y = 'Days survived post-transplant')
```

**Description** :

This shows the same analysis, but considering all the patients without filtering the one who survived.

The scatter plot shows the relationship between the CD34+ cell dose per kilogram of recipient body weight and the post-transplant survival time. Each point represents an individual patient, and the blue line represents a flitted linear regression model with its 95 % confidence interval ( grey area).

A slight positive correlation was observed between CD34+ cell dose per kg and post-transplant survival time, indicating that higher doses may be associated with improved survival. However, the large variance in survival across CD34+ dose levels suggests that this factor alone is insufficient to accurately predicts patient outcomes. Additional clinical and cellular parameters are likely required to better explain survival differences.

```{r}

ggplot(data_clean, 
       aes(x = CD34kgx10d6, 
           y = survival_time)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = TRUE) + 
  labs(title = 'CD34+ dose vs Survival Days',
       x = 'CD34+ cells/kg', 
       y = 'Days survived post-transplant')
```

## 3.2. CD3 ANALYSIS

**Description**:

CD3+ case (without filtering)

```{r}
ggplot(data_clean, 
       aes(x = CD3dkgx10d8, 
           y = survival_time)) +
  geom_point() + 
  geom_smooth(method = 'lm', se = TRUE) + 
  labs(title = 'CD3 dose vs survival Days',
       x = 'CD3+ cells/kg', 
       y = 'Days survived post-transplant')
```

**Description** :

The plot illustrates the relationship between CD3+ dose (cells/kg) and patient survival post-transplant (days). Each black dot represents and individual patient, showing a wide range of survival times (0 to \>3000 days) across spectrum of CD3+ doses. A linear regression (blue line) with its 95% confidence interval (gray band) indicates a very slight negative trend, suggesting that higher CD3+ doses may be weakly associated with shorter survival. The confidence interval broadens at higher doses, reflecting greater uncertainty due to limited data. Overall, there is a substantial variability in survival regardless of CD3+ dose, indicating that CD3+ dose alone is a poor predictor of post-transplant survival. The observed negative association is minimal and likely lacks clinical significance.

```{r}
# Filter only the survival group 
group_surv <- function(data) {
  data_clean |>
    filter(survival_status == '0')
}

# Plot for the filtered data 
ggplot(group_surv(df), 
       aes(x = CD3dkgx10d8, 
           y = survival_time)) +
  geom_point() + 
  geom_smooth(method = 'lm', se = TRUE) + 
  labs(title = 'CD3 dose vs survival Days',
       x = 'CD3+ cells/kg', 
       y = 'Days survived post-transplant')

```

## 3.3. aGvHD III/IV and cGvHD extense ANALYSIS

The development of acute (aGvHD III/IV) and extensive chronic (cGvHD) graft- versus-host diseade has been associated with the dose of CD3+ T cells administered during bone marrow transplantation. Higher CD3+ doses can enhance the graft immune response but also increase the risk of severe GvHD. Therefore, CD3+ dose is critical factor influencing early and late post-transplant immunological outcomes, higlighting the need for a careful balance between benefit and risk

In this analysis, we explore the relationship between CD3+ T cell dose, patient survival time and the development of graft-versus-host disease (GvHD) following bone marrow transplantation. Specifically, we examine how the ocurrence of severe acute GvHD (aGvHD III/IV) and extensive chronic GvHD (cGvHD) correlates with CD3+ cell dose and impacts post-transplant survival. Separate scatter plots are generated for each GvHD type to visualize potential trends and differences, providing insights into the immunological outcomes associated with varying CD3+ doses.

### 3.3.1. aGvHD III/IV (acute Graft-vs-Host Disease)

**Description**:

The analysis of post-transplant survival relative to administered CD3+ dose shows a slight negative trend, suggesting that higher CD3+ doses may be associated with reduced survival, potentially mediated by the development of severe aGvHD. Patients without severe aGvHD are distributed across the full range of CD3+ doses and exhibit highly variable survival times, indicating that the absence of severe aGvHD does not necessarily ensure longer survival. In contrast, patients who developed severe aGvHD tend to cluster at low intermediate CD3+ doses, with some showing shorter survival, consistent with the expected detrimental impact of severe aGvHD. However , the wide dispersion of data points highlights the influence of additional factors, such as patient age, underlying disease, or complications, on survival. Overall, while a weak inverse relationship between CD3+ dose and survival is observed, the patterns is not strictly linear.

```{r}
# Filter only the survival group geom_boxplot() 
group_surv <- function(data_clean) {
  data_clean |>
    filter(survival_status == '0')
}

ggplot(group_surv(data_clean), 
       aes(x = CD3dkgx10d8, 
           y = survival_time, 
           color = factor(aGvHDIIIIV))) + 
  geom_point(size = 3, alpha = 0.7) + 
  scale_color_brewer(palette = 'Set1') + 

  geom_smooth(aes(group = aGvHDIIIIV), method = 'lm', se = FALSE, linetype = 'dashed') +
  theme_minimal() 
```

**Description** :

This analysis examines the impact of infused CD3+ T-cell dose on post-transplant survival in patients with and without extensive chronic graft-versus-host disease (cGvHD). Higher CD3+ doses were associated with reduced survival in both groups, suggesting that excessive donor T-cell infusion may increase harmful immune-mediated toxicity. However, patients who developed extensive cGvHD showed slightly longer survival and a less pronounced decline in response to CD3+ dose, which could indicate a beneficial graft-versus-disease effect.

```{r}

# Filter only the survival group 
group_surv <- function(data_clean) {
  data_clean |>
    filter(survival_status == '0')
}

# Plot

ggplot(group_surv(data_clean),
       aes(x = CD3dkgx10d8, 
           y = survival_time,
           color = factor(extcGvHD))) + 
  geom_point(size = 3, alpha = 0.7) + 
  scale_color_brewer(palette = 'Set1') + 
  labs(
    x = 'Doses of CD3+', 
    y = 'Survival time (days)',
    color = 'cGvHD extense'
  )+
  theme_minimal() +
  geom_smooth(aes(group = extcGvHD), method = 'lm', se = FALSE, linetype = 'dashed')
```

## 4. CMV VIRUS

### 4.1 .Mixed plot between survival and non-survival groupsd:

```{r}

data_new <- data_clean |>
  filter(!is.na(CMVstatus), !is.na(survival_time))

ggplot(data_new, aes(
  x = as.factor(CMVstatus), 
  y = survival_time,
  fill = as.factor(CMVstatus))) + 
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~survival_status) +  # separate between survival and deceasedy pa
  labs(title = "Survival time by CMV compatibility and survival status",
       x = "CMV Compatibility",
       y = "Survival time (days)",
       fill = "CMVstatus") +
  theme_minimal()

```

### 4.2. For the survival group

In patients who survived the transplant, overall survival times were generally prolonged, with medians ranging approximately from 1000 to 1600 days depending on the CMV compatibility between donor and recipient. No clear linear relationship is observed between CMV status and survival, although patients with greater incompatibility show a slight tendency toward higher medians. The data dispersion indicates significant individual variability and the presence of outliers.

```{r}

data_new <- data_clean |>
  filter(!is.na(CMVstatus), !is.na(survival_time)) |>
  filter(survival_status == 0)

ggplot(data_new, aes(
  x = CMVstatus, 
  y = survival_time,
  fill = factor(CMVstatus))) + 
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) + 
  labs(title = 'Survival time by CMV compatibility', 
       x = 'CMV Compatibility', 
       y = 'Survival time (days)') +
  theme_minimal()
```

### 4.3. For the deceased group :

In patients who died after the transplant, survival times were significantly shorter, with medians of approximately 100–200 days, regardless of the level of CMV compatibility. The data distribution suggests that CMV compatibility alone is not a determining factor for survival in this group, and most patients died within the first year post-transplant, with few cases of prolonged survival that could be considered atypical.

```{r}

data_new <- data_clean |>
  filter(!is.na(CMVstatus), !is.na(survival_time)) |>
  filter(survival_status == 1)

ggplot(data_new, aes(
  x = CMVstatus, 
  y = survival_time,
  fill = factor(CMVstatus))) + 
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) + 
  labs(title = 'Survival time by CMV compatibility', 
       x = 'CMV Compatibility', 
       y = 'Survival time (days)') +
  theme_minimal()
```

### 4.4 Donor and Recepient status ( 

## Donor and Recipient status (number of deaths)

The data on CMV matching shows unexpected results:

-   **Highest Risk (58 deaths):** Occurs in CMV-positive recipients **(R+)** receiving a negative donor organ **(D-)**. The R+ patient is the most vulnerable.

-   **Lowest Risk (27 deaths):** Found in the "mismatched" **D+/R-** group. This was the safest combination.

```{r}

table_CMVs <- data_new |>
  group_by(DonorCMV, RecipientCMV) |>
  summarise(count = n(),
            deaths = sum(survival_status == 1),
            survival = sum(survival_status == 0))

ggplot(table_CMVs, aes(x = DonorCMV, y = RecipientCMV, fill = deaths)) +
  geom_tile() +
  geom_text(aes(label = count), color = "white") +
  labs(title = "Donor vs Recipient CMV status (number of patients)",
       x = "Donor CMV", y = "Recipient CMV", fill = "Deaths") +
  theme_minimal()

```
