---
title: "Untitled"
format: html
editor: visual
---

## Clean data.

Takes the bone-marrow.arff and creates a tidy tibble

### Load libraries

```{r}
library("tidyverse")
```

### Clean data

```{r}
data_clean_aug <- read_tsv('data/03_dat_aug.tsv.gz')
```

### Create graph(s) and visual to explain relation with recipient age, disease type, and BMI

### Recipient age

```{r}
 data_clean_aug |>
  select(Recipientgender, Disease, Recipientage, Rbodymass, survival_time, survival_status) |>
  count(Recipientgender, survival_status) |>
   
  ggplot(aes(x = Recipientgender, y = n, fill = factor(survival_status))) +
   
  geom_col(position = "dodge", alpha = 0.5) +
   
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(legend.position = 'bottom')+
  labs( title = 'Survival status among genders',
    x = "Gender",
       y = "Number of survivors/deaths",
       fill = "Survival Status")


```

Note for me: A **factor** in R is a variable made of **categories**, not numbers. We use `factor(survival_status)` so ggplot treats **0 (alive)** and **1 (dead)** as **separate groups** and gives each one a different color.

```{r}
data_clean_aug |>
  ggplot(aes(x = factor(Disease), y = Recipientage, fill = factor(survival_status))) +
  geom_boxplot(alpha = 0.5) + 
   scale_fill_viridis_d() +
  theme_minimal() + 
  labs(
    title = 'Recipient Boxplot',
    x = 'Survival Status', 
    y = 'Age of the recipient',
    fill = 'Survival Status'
  )

  ggsave('Recipient.png', width = 6)


```

This graphs visualizes the survival rate across disease type and ages.

### Disease type

First I'm doing a plot to see the survival rate against the disease type, without gender difference.

```{r}

data_clean_aug |> count(Disease, survival_status, Recipientgender) |>
  ggplot(aes(x = factor(Disease),
             y = n,
             fill = factor(survival_status))) +
  geom_col(position = 'dodge', alpha = 0.75) + 
    scale_fill_viridis_d() +
  theme_minimal() + 
  labs(
    x = 'Survival Status (0 = survivor, 1 = death)', 
    y = 'Number of patients',
    fill = 'Survival Status'
  )


```

Important point –\> none one suffering from lymphoma survived.

### Now I will show the survivors among diseases between genders

```{r}
data_clean_aug |>
  mutate(
    Disease = factor(Disease, 
                     levels = sort(unique(Disease)))
  ) |>
  filter(survival_status =='Alive') |>
  count(Disease, Recipientgender) |>
  ggplot(aes(x = Disease, y = n, fill = Recipientgender)) +
  geom_col(position = "dodge", alpha = 0.5) +
  scale_x_discrete(drop = FALSE) + 
    scale_fill_viridis_d()+
  theme_minimal() +
  labs(
    x = "Type of disease",
    y = "Number of survivors",
    fill = "Gender"
  )

```

### The number of deaths per disease among genders

```{r}
data_clean_aug |>
  mutate(
    Disease = factor(Disease, 
                     levels = sort(unique(Disease)))
  ) |>
  filter(survival_status == 'Dead') |>
  count(Disease, Recipientgender) |>
  ggplot(aes(x = Disease, y = n, fill = Recipientgender)) +
  geom_col(position = "dodge", alpha = 0.5) +
  scale_x_discrete(drop = FALSE) +
    scale_fill_viridis_d()+
  theme_minimal() +
  labs(
    x = "Type of disease",
    y = "Number of deaths",
    fill = "Gender"
  )

```

### Relationship between BMI and survival rate

Graph shows that obese patients had the highest mortality, but also the highest survival rate among the different Rbodymass index values. this time we don't differentiate between women and men, since the indicators used are from WHO and it foes not differ among genders.

```{r}
library(dplyr)
library(ggplot2)

data_clean_aug |>
  mutate(
    Rbodymass = case_when(
      Rbodymass < 18.5 ~ "Underweight",
      Rbodymass >= 18.5 & Rbodymass < 25 ~ "Normal weight",
      Rbodymass >= 25 & Rbodymass < 30 ~ "Overweight",
      Rbodymass >= 30 ~ "Obese"
    ),
    Rbodymass = fct_relevel(Rbodymass,
                                  'Underweight',
                                  'Normal weight',
                                  'Overweight',
                                  'Obese')
  ) |>
  filter(!is.na(Rbodymass)) |>                      
  count(survival_status, Rbodymass) |>
  group_by(Rbodymass) |>
  
  mutate(Rbodymass_percentage = str_c(round(n / sum(n) * 100, 1), "%")) |>
  
  ggplot(aes(x = factor(Rbodymass),
             y = n,
             fill = factor(survival_status))) +
  
  geom_col(position = 'dodge',
           alpha = 0.5) + 
  
  geom_text(aes(label = Rbodymass_percentage),
            position = position_dodge( width = 1),
            vjust = 0,
            size = 3.5)+
  
  theme_minimal() +
  scale_fill_viridis_d() +
  labs(
    title = "Rbodymass vs Survival rate",
    x = "Rbodymass category",   # this is actually BMI group
    y = "Number of recipients",
    fill = "Survival Status"
  ) +
  theme(legend.position = "bottom")

ggsave("Rbodymass.png", width = 6)


```
